CRUSE to Safe Cycling in Ireland

### 2023-09-29 paper.qmd
CRUSE Tool not tool

Uptake of cycling, the main topic of this paper, should be seen in this broader context of transport decarbonisation [@brand2020] and sustainable mobility [@burns2013].
Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
Furthermore, highly efficient e-bikes --- which are seeing rapid uptake in Ireland and other countries --- outperform electric electric cars, which are too heavy and expensive.
Heavy electric cars could slow the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
At the European level, the European Union has a target of reducing greenhouse gas emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050.[^1]

Regarding road traffic casualties, another deadly consequence of inefficient transport systems, the Road Infrastructure Safety Management (RISM) directive (2008/96/EC) requires member states to implement a road safety management system (RSMS) for all public roads.
Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out".[^1]
Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance travelled, the directive requires estimation of distance travelled by mode, down to the road link level.
For active modes, about which there is a paucity of data compared with motorized modes, this is a major challenge.
Better data to inform road safety policies and interventions is a motivation for the CRUSE tool outlined in this paper, and modeling active travel more broadly.

@@

It was in this context that Transport Infrastructure Ireland (TII) commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) project.
Based on the Propensity to Cycle Tool (PCT) for England and Wales, the CRUSE tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.
A key consideration for TII is to ensure both urban and rural Ireland were targeted for cycling infrastructure, as shown in @fig-dublin.
To this end, TII extensively reviewed activity on Irish Roads and how the road network was being used by commuters, cyclists, heavy goods vehicles and other road users across the Irish transport network.
The aim was for the tool to provide strong, national, systematic but locally-specific evidence to monitor cycling friendliness and safety in Ireland, to ensure strategic alignment with national, regional and local policies.
The open source and publicly available nature of the tool ensures more inclusive and evidence-based conversations around cycle network planning between all stakeholders in the transport planning process.
Other use cases in the Irish context includes monitoring the need for existing cycling infrastructure upgrades, undertaking appraisal, as well as safety and exposure information in line with the European Cycling Federation, Dutch Cycling Embassy recommendations and the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
These factors mean that the tool can be seen as an open access 'leverage point' in the planning system [@lovelace2020].

@@

Illustration of the density of route networks generated by the CRUSE tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland.
Source: CRUSE tool, available at <https://cruse.bike/>.
:::

The rest of paper is structured as follows.
In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE tool for Ireland.
In @sec-results, we present the results of the CRUSE tool, including estimates of current cycling levels and future cycling potential at the national, regional and local levels.
In @sec-discussion, we discuss the implications of the results for policy and practice.
In @sec-conclusions, we conclude the paper.

@@

The methods used to generate the evidence presented in the CRUSE tool for Ireland build on the Propensity to Cycle Tool (PCT), which was originally funded by the UK's Department for Transport and developed by a multi-university team.
An important feature of the PCT is that it is open source and publicly available (at [www.pct.bike](https://www.pct.bike/)), allowing its use by all stakeholders in the transport planning process [@lovelace2017].
The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^5] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^6]
.

@@

The CRUSE tool seeks to overcome the following methodological limitations of the original PCT:

@@

However, travel to school and work only constitute a small proportion of all trips.
To overcome this issue, we developed a spatial interaction modelling methodology to estimate the number of trips between each OD pair for additional trip purposes.
The classification of trip purposes used in the CRUSE tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips (not yet implemented).
An overview of the trip purposes used in CRUSE is presented in @fig-trip-purposes.

@@

Trip purposes used in the CRUSE tool.
:::

@@

```{mermaid}
%%| fig-width: 100%
%%| fig-cap: Hypothetical user stories for the CRUSE tool illustrating that it can be used in different ways by different user groups.
%%| label: fig-user-stories
graph LR
    A(Professional user) --> B[CRUSE landing page]
    B --> C[CRUSE national map]
    C --> D[CRUSE county map]
    D --> F[Full county map]
    D --> CS[County statistics]
    D --> Q[Quiet and fast route networks]
    D --> DD[Data downloads]
    DD --> EX{External tools}
    X(Non-professional user) --> Y[CRUSE landing page]
    Y --> Z[CRUSE national map]
    Z --> R[CRUSE county level results]
    Z --> E(Exploration)
    R --> E
    E --> COMS{Communication}
```

@@

Instead of trying to show such use cases, of which there are many hundreds, we present a selection of results to illustrate the main features of the CRUSE tool.

@@

The CRUSE tool provides estimates of cycling potential at the national level, as shown in @fig-national-results.

@@

The CRUSE tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to 'fastest', 'balanced' and 'quietest' route networks.
This provision of multiple scenarios of behaviour changes *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE tool.

@@

The CRUSE tool can help both in terms of describing the current situation, and also prioritise investment in those direct routes with high cycling potential that lack adequate cycling infrastructure, highlighted by @caulfield2012.

@@

RL led the development of the CRUSE tool, with input from JT, EV, HM, EB, PW, GOT, DB, and SM.
SM instigated the project and coordinated its policy impact.
PW, EB and GOT provided project management and transport engineering expertise.

>>>>

Uptake of cycling, the main topic of this paper, should be seen in this broader context of transport decarbonisation [@brand2020] and sustainable mobility [@burns2013].
Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
Furthermore, highly efficient e-bikes --- which are seeing rapid uptake in Ireland and other countries --- outperform electric cars, which are too heavy and expensive.
Over-reliance on electric cars could slow the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
At the European level, the European Union has a target of reducing greenhouse gas emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050.[^1]

Regarding road traffic casualties, another deadly consequence of inefficient transport systems, the Road Infrastructure Safety Management (RISM) directive (2008/96/EC) requires member states to implement a road safety management system (RSMS) for all public roads.
Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out".[^1]
Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance travelled, the directive requires estimation of distance travelled by mode, down to the road link level.
For active modes, about which there is a paucity of data compared with motorized modes, this is a major challenge.
Better data to inform road safety policies and interventions is a motivation for the CRUSE Tool outlined in this paper, and modeling active travel more broadly.

@@

It was in this context that Transport Infrastructure Ireland (TII) commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) project.
Based on the Propensity to Cycle Tool (PCT) for England and Wales, the CRUSE Tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.
A key consideration for TII is to ensure both urban and rural Ireland were targeted for cycling infrastructure, as shown in @fig-dublin.
To this end, TII extensively reviewed activity on Irish Roads and how the road network was being used by commuters, cyclists, heavy goods vehicles and other road users across the Irish transport network.
The aim was for the tool to provide strong, national, systematic but locally-specific evidence to monitor cycling friendliness and safety in Ireland, to ensure strategic alignment with national, regional and local policies.
The open source and publicly available nature of the tool ensures more inclusive and evidence-based conversations around cycle network planning between all stakeholders in the transport planning process.
Other use cases in the Irish context includes monitoring the need for existing cycling infrastructure upgrades, undertaking appraisal, as well as safety and exposure information in line with the European Cycling Federation, Dutch Cycling Embassy recommendations and the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
These factors mean that the tool can be seen as an open access 'leverage point' in the planning system [@lovelace2020].

@@

Illustration of the density of route networks generated by the CRUSE Tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland.
Source: CRUSE Tool, available at <https://cruse.bike/>.
:::

The rest of paper is structured as follows.
In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE Tool for Ireland.
In @sec-results, we present the results of the CRUSE Tool, including estimates of current cycling levels and future cycling potential at the national, regional and local levels.
In @sec-discussion, we discuss the implications of the results for policy and practice.
In @sec-conclusions, we conclude the paper.

@@

The methods used to generate the evidence presented in the CRUSE Tool for Ireland build on the Propensity to Cycle Tool (PCT), which was originally funded by the UK's Department for Transport and developed by a multi-university team.
An important feature of the PCT is that it is open source and publicly available (at [www.pct.bike](https://www.pct.bike/)), allowing its use by all stakeholders in the transport planning process [@lovelace2017].
The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^5] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^6]
.

@@

The CRUSE Tool seeks to overcome the following methodological limitations of the original PCT:

@@

However, travel to school and work only constitute a small proportion of all trips.
To overcome this issue, we developed a spatial interaction modelling methodology to estimate the number of trips between each OD pair for additional trip purposes.
The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips (not yet implemented).
An overview of the trip purposes used in CRUSE is presented in @fig-trip-purposes.

@@

Trip purposes used in the CRUSE Tool.
:::

@@

```{mermaid}
%%| fig-width: 100%
%%| fig-cap: Hypothetical user stories for the CRUSE Tool illustrating that it can be used in different ways by different user groups.
%%| label: fig-user-stories
graph LR
    A(Professional user) --> B[CRUSE landing page]
    B --> C[CRUSE national map]
    C --> D[CRUSE county map]
    D --> F[Full county map]
    D --> CS[County statistics]
    D --> Q[Quiet and fast route networks]
    D --> DD[Data downloads]
    DD --> EX{External tools}
    X(Non-professional user) --> Y[CRUSE landing page]
    Y --> Z[CRUSE national map]
    Z --> R[CRUSE county level results]
    Z --> E(Exploration)
    R --> E
    E --> COMS{Communication}
```

@@

Instead of trying to show such use cases, of which there are many hundreds, we present a selection of results to illustrate the main features of the CRUSE Tool.

@@

The CRUSE Tool provides estimates of cycling potential at the national level, as shown in @fig-national-results.

@@

The CRUSE Tool provides evidence on current and potential future cycling levels across Ireland down to the street level, with potential assigned to 'fastest', 'balanced' and 'quietest' route networks.
This provision of multiple scenarios of behaviour changes *and* multiple scenarios of investment, for example in cycling infrastructure next to major roads vs quiet residential streets, is a key feature of the CRUSE Tool.

@@

The CRUSE Tool can help both in terms of describing the current situation, and also prioritise investment in those direct routes with high cycling potential that lack adequate cycling infrastructure, highlighted by @caulfield2012.

@@

RL led the development of the CRUSE Tool, with input from JT, EV, HM, EB, PW, GOT, DB, and SM.
SM instigated the project and coordinated its policy impact.
PW, EB and GOT provided project management and transport engineering expertise.
